---
title: For loops for quick summaries in R
author: Pete Talbert
date: '2020-02-28'
slug: for-loops-for-quick-summaries-in-r
categories:
  - R
tags:
  - for-loop
  - tidyverse
  - purrr
---

Because I often work with categorical data, I find myself making lots of quick, sorted counts of variables in a dataset. I find that this is a really common technique to get to know a dataset you're working with; I've also noticed David Robinson do it often in his [screencasts](https://www.youtube.com/user/safe4democracy/featured). (If you haven't checked these out, I cannot recommend these enough!)

## Using for loops

As always, I need to make a disclaimer that I know I should be using some type of functional like `lapply` or `purrr::map`, but again since I'm newer to programming, I find it best to make a for loop first to better understand what's happening.

This example uses a sample of the General Social Survey found in the `forcats` package.

```{r warning=FALSE, message=FALSE}
library(tidyverse)

gss <- forcats::gss_cat %>% as_tibble()

gss
```

What I want to do is quick count of the responses for each column of the survey. First, I just try to do it for the first column:

```{r warning=FALSE}
gss %>% 
  group_by(year) %>% 
  summarize(n = n())
```

Easy enough. I could have wrote one less line of code with `count()`, but the reason I am not has to do with how the for loops work. I found that `count()` inside a for loop was nearly impossible, for reasons I have yet to understand.

Now let's write the for loop. As always we want three things: output, sequence, and body.

```{r}
gss_list <- vector("list", ncol(gss)) # 1. output
for (i in 1:ncol(gss)) {              # 2. sequence
  gss_list[[i]] <- gss %>%            # 3. body
    group_by(gss[[i]]) %>% 
    summarize(n = n())
}
#printing the 8th item in the list for an example.
gss_list[[7]]
```

Awesome! That wasn't so bad. And what's nice is, if I print the whole list, I get a nice quick summary of counts for every column in the dataframe.

*Here's the problem:* I want the column name in the dataframe, not `gss[[i]]`, which isn't meaningful. If I had 40 columns for instance, how could I keep track of what's what?

Below, I add another line inside the for loop that replaces `gss[[i]]` in each dataframe in `gss_list` to the original column names.

```{r}
gss_list <- vector("list", ncol(gss)) # 1. output
for (i in 1:ncol(gss)) {              # 2. sequence
  gss_list[[i]] <- gss %>%            # 3. body
    group_by(gss[[i]]) %>% 
    summarize(n = n()) %>% 
    ungroup()
  colnames(gss_list[[i]])[1] <- names(gss)[i] #here I rename the column to its orig name.
}
#printing the 8th item in the list for an example.
gss_list[[7]]
```

It's not the prettiest code, but it does the trick. I'd love a more elegant solution, but for now, it works.

## A quick plot

I always end by doing a quick plot, because really what's the point of summarizing data like this without visualizing it in some way?

First, though, I want to group by the first column `year` for all my counts, so I'll tweak the for loop again.

```{r}
gss_list <- vector("list", ncol(gss)) # 1. output
for (i in 1:ncol(gss)) {              # 2. sequence
  gss_list[[i]] <- gss %>%            # 3. body
    group_by(year, gss[[i]]) %>% #adding the year column
    summarize(n = n()) %>% 
    ungroup()
  colnames(gss_list[[i]])[2] <- names(gss)[i] #here I rename the column to its orig name. Note: it's the second column now!
}
#printing the 8th item in the list for an example.
gss_list[[7]]
```

Okay, now on to plotting! This is also a chance to show off Julia Silge's awesome `reorder_within()` [function](https://juliasilge.com/blog/reorder-within/) that allows you to easily reorder factors within each facet using `facet_wrap()`.

```{r warning=FALSE, message=FALSE, fig.height=8}
library(scales)
library(tidytext) #this has reorder_within() along with a lot of great functions for working with text.

theme_set(theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"),
        axis.text = element_text(size = 8))
)

gss_list[[7]] %>% 
  mutate(relig = reorder_within(relig, n, year)) %>% #use this inside of mutate
  ggplot(aes(x = relig, y = n, fill = relig)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~year, scales = "free_y") +
  scale_x_reordered() + #and this to scale it properly.
  scale_fill_viridis_d(direction = -1) +
  scale_y_log10(labels = comma_format()) +
  theme(legend.position = "none") +
  labs(
    title = "Although the top five have remained steady,\nthere's been lots of movement in minority religions.",
    subtitle = "Count of religious affiliation, 2000-2014.",
    x = element_blank(),
    y = element_blank(),
    caption = "A sample of categorical variables from the General Social survey."
  )
  
```

Another thing to note that David Robinson got me hooked on is `scale_y_log10()`; without it in this particular plot, it would be difficult to see how the smaller minority religions have changed across time.







