---
title: Pivoting Data in SQL vs. R
author: ''
date: '2018-11-22'
slug: pivoting-data-in-sql-vs-r
categories:
  - SQL
  - R
tags:
  - ms-sql-server
  - pivot
  - tidyverse
  - spread
---

Often with education data, I need to pivot data to go from long format to wide. For instance, I often have test scores for different test subjects of a single assessment. For one student, I will have to rows, one for one subject and one for another. This also happens for strands and domains of a test.

```{r  message=FALSE}
library(tidyverse)
library(DBI)
library(odbc)
library(knitr)

con <- dbConnect(odbc::odbc(), "nicodemus")

query <- dbSendQuery(con, "select * from Demo.dbo.tbl_testScore")
testScore <- dbFetch(query)

testScore <- as_tibble(testScore)

testScore
```

Above is a very simple table of test scores where `personID` is the unique identifier, and records duplicate by `testSubject` for each student. How would I break out `testSubject` as columns and put the `scaleScore` in the values of those columns? Like say then to plot on a scatterplot?

## Pivoting in SQL

In SQL, the syntax is the following. It's important to note (as opposed to the `dplyr` option later) that this is an *aggregate* function. So the `pivot` function requires some aggregate like `MAX()`, `MIN()`, `SUM()`, or `AVG()`. You choose the column to pivot out and then the values you want to place in those new columns given some aggregation. Since I know I only have one record for each student + test subject combination, I chose `MAX()` to just give me that one score.

```{sql connection = con, output.var = "testScorePiv"}
select *
from Demo.dbo.tbl_testScore
pivot(
  MAX(scaleScore) for testSubject in ([Math],[Reading]) --aggregate function of one column for another column, listing the unique possibilities in brackets.
) piv
```

```{r, echo = FALSE}
as_tibble(testScorePiv)
```

There you have it! Pretty straight-forward. One major downside to pivoting in SQL is that you need to know exactly the unique values of the pivoting column. In our example, I had to hardcode `Math` and `Reading` in the brackets, and I knew that this test only has these two subjects. But what if the pivoting column has an unknown number of values, or the data changes?

It makes sense that a database programming language wouldn't have the function to automatically pivot out all the values, as RDBMS's are highly structed.

## Pivoting in R with dplyr

In the `dplyr` package you can use the `spread()` and `gather()` function to pivot and unpivot columns respectively. `spread()` requires that you put the pivoted column in `key` and the values you want to place in the pivoted columns in `value`.

```{r}
testScore %>% 
  spread(key = testSubject, value = scaleScore)
```

Same result as above with SQL. What I struggled with the most initially with this function (coming from the SQL world) was that there must be only one value to put in the pivoted columns; otherwise, it erors. The SQL solution, on the other hand, has you put an aggregate function, so that if it meets two values, it has an aggregation to get to one.

### What if you don't have unique values to pivot in R?

Let's say in our example that a student had *two* Math and Reading scores and you needed to take the maximum for each subject. See the table below where the last student has two more records.

```{r}
testScoreMore <- testScore %>% 
  add_row(personID = 1007, testName = "MCA-III", testSubject = "Math", scaleScore = 363) %>%
  add_row(personID = 1007, testName = "MCA-III", testSubject = "Reading", scaleScore = 395)
```

```{r, echo = FALSE}
as_tibble(testScoreMore)
```

If I try to run the same code as above it gives an error:

```{r error = TRUE}
testScoreMore %>% 
  spread(key = testSubject, value = scaleScore)
```

I think the easiest solution is to just do your aggregation with `group_by()` and `summarize()` earlier in the pipeline, and then do your `spread()` after.

```{r}
testScoreMore %>% 
  group_by(personID, testName, testSubject) %>% #get to unique identifiers with these two lines.
  summarize(scaleScore =  max(scaleScore)) %>% #using max here.
  spread(key = testSubject, value = scaleScore)
```

Cool! Hey, now that we have columns for each subject, is there a relationship between them?

```{r}
testScoreMore %>% 
  group_by(personID, testName, testSubject) %>% 
  summarize(scaleScore =  max(scaleScore)) %>% 
  spread(key = testSubject, value = scaleScore) %>% 
  ggplot(aes(x = Math, y = Reading)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

